---
title: "Playing with GIS files in R"
output: html_notebook
---

This document describes how to import GIS files into R and create pretty maps. There are many useful packages out there for doing spatial data manipulation and analyses. I only know a few. 

Some packages that I use for data manipulations and plotting:

```{r}
library(tidyverse)   # for data manipulation
library(ggplot2)     # for plotting
library(sf)          # spatial data manipulation
library(lubridate)   # dealing with date and time

```

GIS data were obtained from here:
https://geodata.dep.state.fl.us/
https://shoreline.noaa.gov/data/datasheets/usgs.html

I unzipped downloaded files and moved the resulting folders into this project folder with a new folder called "shapefiles." 

```{r}
# To bring in shapefiles into the workspace, we specify the folder that contains GIS files
# each GIS object comes with multiple files but shares one name.
# If one folder contains multiple objects, specify it in the 
# layer argument:
# FL baseline from fl_zip on the Gulf of Mexico side
FL_baseline_GofM <- st_read("shapefiles/fl_zip_GofM", layer = "fl_baseline")
FL_1998_2001_GofM  <- st_read("shapefiles/fl_zip_GofM", layer = "fl1998_2001")

#FL_transects_lt_GofM  <- st_read("shapefiles/fl_zip_GofM", layer = "fl_transects_lt")

# FL baseline from fl_zip on the Atlantic side
FL_baseline_Atl <- st_read("shapefiles/fl_zip_Atl", layer = "fl_baseline")
FL_1999_Atl  <- st_read("shapefiles/fl_zip_Atl", layer = "fl1999")

#FL_transects_lt_Atl  <- st_read("shapefiles/fl_zip_Atl", layer = "fl_transects_lt")

# take a look at it
p.1 <- ggplot() +
  geom_sf(data = FL_baseline_GofM) +
  geom_sf(data = FL_baseline_Atl) +
  geom_sf(data = FL_1998_2001_GofM) + 
  geom_sf(data = FL_1999_Atl)
p.1

```

Not a very interesting plot...

```{r}
# They misspelled "Research" with no second e... Took me a while to figure out why it was returning an error
FL_NERR <- st_read("shapefiles/FL_NERR", 
                  layer = "Florida_National_Estuarine_Resarch_Reserves__NERR__Boundaries")
                    
                    #"Florida_National_Estuarine_Research_Reserves__NERR__Boundaries")

FL_StParks <- st_read("shapefiles/Florida_State_Parks",
                      layer = "Florida_State_Parks_Boundaries")

Coastal_Inlets <- st_read("shapefiles/Coastal_Inlets_Management",
                          layer = "Coastal_Inlets_Management")

p.2 <- ggplot() +
  geom_sf(data = FL_baseline_GofM) +
  geom_sf(data = FL_baseline_Atl) +
  geom_sf(data = FL_NERR, color = "red") +
  geom_sf(data = FL_StParks, color = "green") +
  geom_sf(data = Coastal_Inlets, color = "orange")


p.2
```

Need more examples ... 

