---
title: "Playing with GIS files in R"
output: html_notebook
---

This document describes how to import GIS files into R and create pretty maps. There are many useful packages out there for doing spatial data manipulation and analyses. I only know a few. 

Some packages that I use for data manipulations and plotting:

```{r}
library(tidyverse)   # for data manipulation
library(ggplot2)     # for plotting
library(sf)          # spatial data manipulation
library(lubridate)   # dealing with date and time
library(readr)

```

GIS data were obtained from here:
https://geodata.dep.state.fl.us/
https://shoreline.noaa.gov/data/datasheets/usgs.html

I unzipped downloaded files and moved the resulting folders into this project folder with a new folder called "shapefiles." 

```{r}
# To bring in shapefiles into the workspace, we specify the folder that contains GIS files
# each GIS object comes with multiple files but shares one name.
# If one folder contains multiple objects, specify it in the 
# layer argument:
# FL baseline from fl_zip on the Gulf of Mexico side
FL_baseline_GofM <- st_read("shapefiles/fl_zip_GofM", layer = "fl_baseline")
FL_1998_2001_GofM  <- st_read("shapefiles/fl_zip_GofM", layer = "fl1998_2001")

#FL_transects_lt_GofM  <- st_read("shapefiles/fl_zip_GofM", layer = "fl_transects_lt")

# FL baseline from fl_zip on the Atlantic side
FL_baseline_Atl <- st_read("shapefiles/fl_zip_Atl", layer = "fl_baseline")
FL_1999_Atl  <- st_read("shapefiles/fl_zip_Atl", layer = "fl1999")

#FL_transects_lt_Atl  <- st_read("shapefiles/fl_zip_Atl", layer = "fl_transects_lt")

# take a look at it
p.1 <- ggplot() +
  geom_sf(data = FL_baseline_GofM) +
  geom_sf(data = FL_baseline_Atl) +
  geom_sf(data = FL_1998_2001_GofM) + 
  geom_sf(data = FL_1999_Atl)
p.1

```

Not a very interesting plot...

```{r}
# They misspelled "Research" with no second e... Took me a while to figure out why it was returning an error
FL_NERR <- st_read("shapefiles/FL_NERR", 
                  layer = "Florida_National_Estuarine_Resarch_Reserves__NERR__Boundaries")
                    
                    #"Florida_National_Estuarine_Research_Reserves__NERR__Boundaries")

FL_StParks <- st_read("shapefiles/Florida_State_Parks",
                      layer = "Florida_State_Parks_Boundaries")

Coastal_Inlets <- st_read("shapefiles/Coastal_Inlets_Management",
                          layer = "Coastal_Inlets_Management")

p.2 <- ggplot() +
  geom_sf(data = FL_baseline_GofM) +
  geom_sf(data = FL_baseline_Atl) +
  geom_sf(data = FL_NERR, color = "red") +
  geom_sf(data = FL_StParks, color = "green") +
  geom_sf(data = Coastal_Inlets, color = "orange")


p.2
```


How do we add our own data to these plots?

Save the original text data file into the data sub directory within the project folder. 

Then, bring in the data. I use read_csv in the readr package, which requires you to define the type of each column. This may seem a bit of extra work but it's good to make sure what's coming in from a data file is what you expect. This is one way to check for errors in data. 

```{r}

col_defs <- cols(location_name = col_character(),
                 site_name = col_character(),
                 trip_code = col_character(),
                 trip_year = col_integer(),
                 reef_name = col_character(),
                 reef_type = col_character(),
                 set_code = col_character(),
                 maxn = col_integer(),
                 common_name = col_character(),
                 genus = col_character(),
                 species = col_character(),
                 species_group = col_character(),
                 family = col_character(),
                 event_time_mil = col_integer(),
                 event_time_mins = col_character(),
                 set_lat = col_double(),
                 set_long = col_double(),
                 visibility = col_integer(),
                 depth = col_double(),
                 bait = col_character(),
                 current_flow_estimated = col_character())

BRUVS_data <- read_csv("data/FLkeys_maxn_obs.csv", 
                       col_types = col_defs)
```

We then convert these data to spatial points. 

The sf package provides easy ways to accomplish all these things. Take a look at their help files, which can be found here: https://r-spatial.github.io/sf/articles/sf1.html

Also, here is a similar question asked to StackOverflow: https://stackoverflow.com/questions/29736577/how-to-convert-data-frame-to-spatial-coordinates

crs (coordinate reference system) needs to be figured out - I assume this was WGS84, which is the most common one when using GPS to record lat/lon. 
```{r}

BRUVS_sf <- st_as_sf(x = BRUVS_data, 
                     coords = c("set_long", "set_lat"),
                     crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

str(BRUVS_sf)

```


```{r}
p.3 <- p.2 + 
  geom_sf(data = BRUVS_sf,
          aes(color = as.factor(trip_year)))

p.3
```

We can extract some specific species...

```{r}
BRUVS_sf %>% filter(genus == "Sphyrna") -> BRUVS_Sphyrna_sf

p.4 <- p.2 + 
  geom_sf(data = BRUVS_Sphyrna_sf,
          aes(color = as.factor(trip_year), size = maxn))
p.4
```

We really don't need the entire FL for these plots... and add maxN to the plot and change the legend titles. 

```{r}

p.4 + coord_sf(xlim = c(-82, -80), ylim = c(24, 25.5)) +
  labs(title = "Sphyrna") + 
  labs(color = "Year", size = "maxN")

```



